---
title: "Major Definition"
author: "Miao Li"
date: "`r format(Sys.time(), '%d %B, %Y')`"
#toc: true
format: 
  html:
    self-contained: true
    #output-file: "majors/major_definition"
    #output-ext:  "html"
tbl-colwidths: [15, 75, 25]
editor: visual
execute: 
  echo: FALSE
  message: FALSE
  warning: false
---

## This file documents how majors were coded for the MIDFIELD data.

```{r read_data, message=FALSE, warning=FALSE}
library(data.table)
library(readr)
library(tidyverse)
library(knitr)

major <- read_csv("D:/OneDrive - Clemson University/My Grant Project/Clemson_projects/MIDFIELD/Data/R/major_in_out.csv")
major <- setDT(major)
cip2name <- unique(major[,.(CIP2matri, CIP2matriName)][order(CIP2matri)])
cip4name <- unique(major[,.(CIP4matri, CIP4matriName)][order(CIP4matri)])

```

## STEM

The following majors were coded as STEM majors:

```{r STEM}
stemmajorct <- major %>% filter(CIP2matri %chin% c("11", "14", "15", "25", "26", "27", "40", "41")) %>% 
  count(CIP4matri)
stemcode <- cip4name[stemmajorct, on=c('CIP4matri')]
names(stemcode) <- c('CIP4', 'CIP4Name', 'Count')
kable(stemcode, caption = 'STEM Majors')  
```

## Social Sciences [^1]

[^1]: In an alternative coding approach, Economics were grouped together with Business.

The following majors were coded as Social Sciences:

```{r Social_Sciences}
ssmajorct <- major %>% filter(CIP2matri %chin% c("09", "13", "19", "22", "31", "42", "43", "44", "45")) %>% 
  count(CIP4matri)
sscode <- cip4name[ssmajorct, on=c('CIP4matri')]
names(sscode) <- c('CIP4', 'CIP4Name', 'Count')
kable(sscode, caption = 'Social Science Majors')
```

## Humanities

The following majors were coded as Humanities:

```{r Humanities}
hmajorct <- major %>% filter(CIP2matri %chin% c("05", "16", "23", "24", "33", "38", "39", "54")) %>% 
  count(CIP4matri)
hcode <- cip4name[hmajorct, on=c('CIP4matri')]
names(hcode) <- c('CIP4', 'CIP4Name', 'Count')
kable(hcode, caption = 'Humanities Majors')

```

## Business

The following majors were coded as Business:

```{r Business}
bct <- major %>% filter(CIP2matri %chin% c("52")) %>% 
  count(CIP4matri)
bcode <- cip4name[bct, on=c('CIP4matri')]
names(bcode) <- c('CIP4', 'CIP4Name', 'Count')
kable(bcode, caption = 'Business Majors')

```

## Health Professions

The following majors were coded as Health Professions:

```{r Health_Professions}
 
hpct <- major %>% filter(CIP2matri %chin% c("51")) %>% 
  count(CIP4matri)
hpcode <- cip4name[hpct, on=c('CIP4matri')]
names(hpcode) <- c('CIP4', 'CIP4Name', 'Count')
kable(hpcode, caption = 'Health Professions Majors')

```

## Agriculture

The following majors were coded as Agriculture:

```{r Agriculture}
agct <- major %>% filter(CIP2matri %chin% c("01", "03")) %>% 
  count(CIP4matri)
agcode <- cip4name[agct, on=c('CIP4matri')]
names(agcode) <- c('CIP4', 'CIP4Name', 'Count')
kable(agcode, caption = 'Agriculture Majors')

```

## Other

The following majors were coded as Other:

```{r Other}
otct <- major %>% filter(CIP2matri %chin% c("04", "28", "30", "33", "46", "47", "48", "49", "50")) %>% 
  count(CIP4matri)
otcode <- cip4name[otct, on=c('CIP4matri')]
names(otcode) <- c('CIP4', 'CIP4Name', 'Count')
kable(otcode, caption = 'Other Majors')

```
